Quantile Trend Filtering via Spingarn's Method

### 2017-06-01 Notes/detrendify_spingarn.tex
Added some more references and version 2 of the algorithm.

Examples:
\begin{itemize}
\item Air quality 
\item ECG \cite{Sanyal2012}
\item Chromatogram baseline estimation  \cite{Ning2014, Ilewicz2015}
\item Galaxy spectrum baseline estimation \cite{Ilewicz2015,Bacher2016}
\end{itemize}

@@

\begin{eqnarray*}
\V{x}^+ & = & \prox{t\psi}{\V{z}} \\
\V{y}^+ & = & P_V(2\V{x}^+ - \V{z}) \\
\V{z}^+ & = & \V{z} + \V{y}^+ - \V{x}^+.
\end{eqnarray*}
The parameter $t$ is a step-size which can be fixed at 1 to guarantee convergence. The mapping $P_V$ is the orthogonal projection onto the set $V$.

@@

If we set $\psi(\V{\theta}, \V{\eta}) = f_1(\V{\theta}) + f_2(\V{\eta})$ and $V = \{ \V{z}\Tra = (\V{\theta}\Tra, \V{\eta}\Tra) : \V{\eta} = \M{D}\V{\theta} \}$, then we can apply Spingarn's method. Note that
\begin{eqnarray*}
\prox{th}{\V{\theta}, \V{\eta}} & = & (\prox{tf_1}{\V{\theta}}, \prox{tf_2}{\V{\eta}}) \\
P_V(\V{\theta}, \V{\eta}) & = & \begin{pmatrix}
\M{I} \\ \M{D}
\end{pmatrix}
\begin{pmatrix}
\M{I} + \M{D}\Tra\M{D}
\end{pmatrix}\Inv
\begin{pmatrix}
\V{\theta} + \M{D}\Tra\V{\eta}
\end{pmatrix},
\end{eqnarray*}
where the projection $P_V$ requires a banded linear system solve, with bandwidth $k + 1$.
This linear solve can be accomplished in $\mathcal{O}(n(k + 1)^2)$. The first solve using a banded Cholesky decomposition requires $\mathcal{O}(n(k + 1)^2)$. Subsequent solves require
$\mathcal{O}(n(k+1))$.

We should also consider trying the reparameterization for ADMM trend filtering proposed in \cite{Ramdas2016}.

@@

\begin{itemize}
\item The overall computational complexity is essentially linear $\mathcal{O}(nk^2)$ for the initial banded Cholesky decomposition and the per-iteration complexity is $\mathcal{O}(nk)$.
\item One could also apply Anderson acceleration to reduce the number of Spingarin updates, since the Douglas-Rachford algorithm is a fixed point algorithm.
\end{itemize}

>>>>

Examples:
\begin{itemize}
\item Air quality 
\item ECG \cite{Sanyal2012}
\item Chromatogram baseline estimation  \cite{Ning2014, Ilewicz2015}
\item Galaxy spectrum baseline estimation \cite{Ilewicz2015,Bacher2016}
\item Identifying absorption dips from black body radiation
\end{itemize}

@@

\begin{eqnarray*}
\Vn{x}{k+1} & = & \prox{t\psi}{\Vn{z}{k}} \\
\Vn{y}{k+1} & = & P_V\left(2\Vn{x}{k+1} - \Vn{z}{k} \right) \\
\Vn{z}{k+1} & = & \Vn{z}{k} + \lambda^{(k)}\left(\Vn{y}{k+1} - \Vn{x}{k+1} \right).
\end{eqnarray*}
The parameter $t$ is a step-size and $\lambda^{(k)}$ is Krasnosel'ski{\u i}-Mann iteration (Need citation). We require that $\lambda^{(k)} \in ]0, 2[$ and that $\sum_n \lambda^{(k)} (2-\lambda^{(k)}) = \infty$. The mapping $P_V$ is the orthogonal projection onto the set $V$.  Note that the algorithm iterates $\Vn{x}{k}$ will converge to a solution to problem $\Eqn{spingarn}$ \cite{Combettes2005}. We need to experiment with different over / under-relaxation parameters $\lambda^{(k)}$ and step sizes $t$. It will converge if we take $\lambda^{(k)} = 1$ and $t = 1$, but we may be able to converge faster in practice by taking non-trivial values.

@@

If we set $\psi(\V{\theta}, \V{\eta}) = f_1(\V{\theta}) + f_2(\V{\eta})$ and $V = \{ \V{z}\Tra = (\V{\theta}\Tra, \V{\eta}\Tra) : \V{\eta} = \M{D}\V{\theta} \}$, then we can apply Spingarn's method. Note that
\begin{eqnarray*}
\prox{th}{\V{\theta}, \V{\eta}} & = & (\prox{tf_1}{\V{\theta}}, \prox{tf_2}{\V{\eta}}) \\
P_V(\V{\theta}, \V{\eta}) & = & \begin{pmatrix}
\M{I} \\ \M{D}
\end{pmatrix}
\begin{pmatrix}
\M{I} + \M{D}\Tra\M{D}
\end{pmatrix}\Inv
\begin{pmatrix}
\V{\theta} + \M{D}\Tra\V{\eta}
\end{pmatrix},
\end{eqnarray*}
where the projection $P_V$ requires a banded linear system solve, with bandwidth $k + 1$.
This linear solve can be accomplished in $\mathcal{O}(n(k + 1)^2)$. The first solve using a banded Cholesky decomposition requires $\mathcal{O}(n(k + 1)^2)$. Subsequent solves require
$\mathcal{O}(n(k+1))$. We can use RccpArmadillo to do banded Cholesky \cite{Eddelbuettel2014}.

@@

\begin{itemize}
\item The overall computational complexity is essentially linear $\mathcal{O}(nk^2)$ for the initial banded Cholesky decomposition and the per-iteration complexity is $\mathcal{O}(nk)$.
\item One could also apply Anderson acceleration \cite{Walker2011} to reduce the number of Spingarn updates, since the Douglas-Rachford algorithm is a fixed point algorithm.
\end{itemize}

%% ----------------------------------------------------------------------
%% Applying Spingarn's Method to Quantile Trend Filtering version 2
%% ----------------------------------------------------------------------
\section{Applying Spingarn's Method to Quantile Trend Filtering version 2}

Following the specialized ADMM algorithm for trend filtering \cite{Ramdas2016}, we can also take advantage of fast exact solvers of the one-dimensional fussed lasso problem \cite{davies2001, Johnson2013}. The first method is based on taut strings, and the second is based on dynamic programming. Both of these methods run in linear time. A third exact method with worst case quadratic penalty but linear time in typical cases was proposed in \cite{Condat2013}. C code is available for all three methods. We should compare them.

What is the reparameterization?

\begin{eqnarray*}
\text{minimize}\; && f_1(\V{\theta}) + f_2(\V{\eta}) \\
\text{subject to}\; && \V{\eta} = \M{D}\V{\theta}
\end{eqnarray*}
where
\begin{eqnarray*}
f_1(\V{\theta}) = \frac{1}{n}\sum_{i=1}^n \rho_\tau(\VE{y}{i} - \VE{\theta}{i}) \quad\quad \text{and} \quad\quad
f_2(\V{\eta}) = \lambda\lVert \Mn{D}{1}\V{\eta} \rVert_1.
\end{eqnarray*}

Difference with version 1: If we are interested in a 3rd order penalization, then $\M{D} = \Mn{D}{2}$ in version 2 as opposed to $\M{D} = \Mn{D}{3}$ in version 1. The proximal mapping for $f_2$ is solved using one of the exact solvers \cite{davies2001, Johnson2013, Condat2013}. Show some timing results between the different versions. Point readers to \cite{Ramdas2016} for a discussion of why this minor reparameterization may lead to speed up.

\section{Other Practical Issues}

\begin{itemize}
	\item Read compressed data
	\item May want to read data directly in C and not pull into R; make R just an interface
	\item Use historical data to choose $\tau$ and $\lambda$
	\item 
\end{itemize}
